---
title: "stchlk_exome_file_load.Rmd"
author: "Kevin Stachelek"
date: "1/30/2018"
output: html_document
---

```{r m2_file_load_00, eval = F}
source("related_functions.R")

```

load rb datasets. Keep in mind that only samples c("24-T", "28-C", "28-T", "29-T", "29-C", "31-T", "31-C", "41-C", "46-T", "46-C", "48-T", "48-C") have ready rb variants detected 

# 1. load packages

```{r m2_file_load01, warning=FALSE, include = F}
library(tidyverse)
library(stchlkExome)
suppressMessages(library(EnsDb.Hsapiens.v86))
edb <- EnsDb.Hsapiens.v86
library(org.Hs.eg.db)
library(ggrepel)
library(S4Vectors)
library(ensemblVEP)
library(plyranges)
library(glue)
library(data.table)
library(rprojroot)
library(fs)
library(GenVisR)
library(gridExtra)
library(WebGestaltR)
library(ggridges)

proj_dir = proj_dir = rprojroot::find_root(criterion = has_file_pattern("*.Rproj"))

sample_types <- c("tumors", "cell_lines", "normals", "reynolds")

```

# 2. vcf filenames

```{r m2_file_load02}
#clinvar <-  read_tsv('../Homo_sapiens/ClinVar/clinvar_alleles.single.b37.tsv',sep='\t',header=T,quote='',comment.char='')

# load input --------------------------------------------------------------
# variants w/o vc filtering 
m2_tumor_unfiltered_filenames <- list.files(path="~/rb_pipeline/output/mutect2", pattern="^[[:digit:]]{2}-T_1_mutect2_vep.vcf$", full.names = TRUE)
m2_cell_line_unfiltered_filenames <- list.files(path="~/rb_pipeline/output/mutect2", pattern="^[[:digit:]]{2}-CL_1_mutect2_vep.vcf$", full.names = TRUE)
m2_pon_unfiltered_filenames <- list.files(path="~/rb_pipeline/output/mutect2", pattern="^[[:digit:]]{2}-N_1_mutect2_vep.vcf$", full.names = TRUE)
m2_reynolds_unfiltered_filenames <- list.files(path="~/rb_pipeline/output/mutect2", pattern="^[[:digit:]]{3}-CL_1_mutect2_vep.vcf$", full.names = TRUE)
```



# 3. read mutect2 vcf files

```{r m2_file_load03b}
# unfiltered_filenames <- list(m2_tumor_unfiltered_filenames, 
#                   m2_cell_line_unfiltered_filenames,
#                   m2_pon_unfiltered_filenames,
#                   m2_reynolds_unfiltered_filenames)
# 
# unfiltered_sample_types <- list("m2_tumor_unfiltered",
#                      "m2_cell_line_unfiltered",
#                      "m2_normal_unfiltered",
#                      "m2_reynolds_unfiltered")

unfiltered_filenames <- list(m2_tumor_unfiltered_filenames, 
                  m2_cell_line_unfiltered_filenames)

unfiltered_sample_types <- list("m2_tumor_unfiltered",
                     "m2_cell_line_unfiltered")

# load and save vcflists
vcf_lists <- purrr::map2(unfiltered_filenames, unfiltered_sample_types, load_vcfs)
```


# 4. load m2 vcf rds files

```{r m2_file_load04b}
unfiltered_vcf_lists <- paste0("~/rb_pipeline/results/SNV/m2_", c("cell_line", "tumor", "normal", "reynolds"), "_unfiltered_list.rds") %>% 
  purrr::map(readRDS)



```

# 5. collate m2 vcfs to tibble

```{r m2_file_load05}
tidy_functions <- list(mutect2_tn_tidy, mutect2_tn_tidy, mutect2_pon_tidy, mutect2_pon_tidy)
m2_types <- list("tn", "tn", "pon", "pon")

unfiltered_vcf_lists <- purrr::map2(unfiltered_vcf_lists, m2_types, ~structure(.x, class = .y))

m2_tidy_samples_unfiltered <- purrr::map2(unfiltered_vcf_lists, tidy_functions, collate_vcfs)
```

# 6. save collated m2 vcfs

```{r m2_file_load05b}
sample_types <- c("tumors", "cell_lines", "normals", "reynolds")

names(m2_tidy_samples_unfiltered) <- sample_types

# save raw variant caller files------------------------------
m2_collated_vars_ps <- fs::path(proj_dir, "results", "SNV", paste0("tidy_m2_", sample_types, "_unfiltered.rds"))

map2(m2_tidy_samples_unfiltered, m2_collated_vars_ps, saveRDS)

```

# 6. load unrefined m2 variants

```{r m2_file_load06b}
m2_collated_vars_ps_unfiltered <- fs::path(proj_dir, "results", "SNV", paste0("tidy_m2_", sample_types, "_unfiltered.rds"))

filter_vars <- function(vars){
  vars <- vars %>% 
    dplyr::filter(FILTER %in% c(".", "PASS")) %>% 
    dplyr::filter(!is.na(HGVSc)) %>% 
    identity()
}

m2_collated_vars_unfiltered <- 
  m2_collated_vars_ps_unfiltered %>% 
  purrr::map(readRDS) %>% 
  # purrr::map(filter_vars) %>%
  purrr::set_names(sample_types) %>% 
  identity()

purrr::map2(c("m2_vc_tumors", "m2_vc_cell_lines", "m2_vc_reynolds"), m2_collated_vars_unfiltered, print_tl)

```

# 7. panel of normals and variant caller filter m2 variants

```{r m2_file_load07}

m2_tidy_samples <- m2_collated_vars_unfiltered

m2_tidy_samples$reynolds <- dplyr::rename(m2_tidy_samples$reynolds, GT.TUMOR = GT)

reference_samples <- m2_tidy_samples[c("tumors", "cell_lines", "reynolds")]

# anti join panel of normals variants
m2_tidy_samples <- purrr::map(reference_samples, pon_subtract_m2_vars, m2_tidy_samples[["normals"]])

m2_somatic_vars_paths <- paste0("~/rb_pipeline/results/SNV/mutect2_variants_wo_pon_", c("tumors", "cell_lines", "reynolds"), "_unfiltered.rds")

purrr::map2(m2_tidy_samples, m2_somatic_vars_paths, saveRDS)

```


# 8. load pon subtracted mutect2 variants

### filtered

```{r m2_file_load08}

sample_types <- c("tumors", "cell_lines", "reynolds")

m2_somatic_vars_paths <- paste0("~/rb_pipeline/results/SNV/mutect2_variants_wo_pon_", sample_types, ".rds") %>% 
  set_names(sample_types)

m2_tidy_samples <- purrr::map(m2_somatic_vars_paths, readRDS)

# purrr::map2(c("tumors", "cell_lines", "reynolds"), m2_tidy_samples, print_tl)
```

### unfiltered

```{r m2_file_load08}

sample_types <- c("tumors", "cell_lines", "reynolds")

m2_somatic_vars_paths <- paste0("~/rb_pipeline/results/SNV/mutect2_variants_wo_pon_", sample_types, "_unfiltered.rds") %>% 
  set_names(sample_types)

m2_tidy_samples_unfiltered <- purrr::map(m2_somatic_vars_paths, readRDS)

# purrr::map2(c("tumors", "cell_lines", "reynolds"), m2_tidy_samples, print_tl)
```


# 9. refine and create table of all mutect2 somatic variants

### filtered

```{r m2_file_load09}

# select columns to enable clean rbinding of mutect2 variants

m2_tidy_tumors <- m2_tidy_samples$tumors

m2_tidy_cell_lines <- dplyr::select(m2_tidy_samples$cell_lines, colnames(m2_tidy_samples$tumors)) %>% 
  identity()

min_colnames <- colnames(m2_tidy_samples$tumors)[!(grepl("NORMAL", colnames(m2_tidy_samples$tumors)))]

min_colnames <- c(min_colnames, "GT", "AF", "AD.1", "AD.2")

m2_tidy_reynolds <- dplyr::select(m2_tidy_samples$reynolds, one_of(min_colnames)) %>% 
  identity()
```


```{r }

# filter mutect2 variants
m2_tidy_samples <- list(m2_tumor = m2_tidy_tumors, m2_cl = m2_tidy_cell_lines, m2_reynolds = m2_tidy_reynolds)

m2_tidy_samples <- 
  m2_tidy_samples %>% 
  purrr::map(~dplyr::filter(.x, FILTER %in% c("PASS", "."))) %>% 
  purrr::map(~dplyr::select(.x, gnomAD_AF = one_of("gnomAD_AF", "gnomad.AF"), everything()))

m2_tidy_samples <- purrr::map2(m2_tidy_samples, c("tn", "tn", "pon"), filter_variant_set) %>% 
  set_names(c("vc_T", "vc_CL", "reynolds_CL"))

m2_tidy_samples$reynolds_CL <- dplyr::mutate(m2_tidy_samples$reynolds_CL, AF.TUMOR = AF)

m2_filt_variant_paths <- paste0("~/rb_pipeline/results/SNV/", c("clean_vc_tumor_variants.txt", "clean_vc_cell_line_variants.txt", "clean_reynolds_cell_line_variants.txt"))

purrr::map2(m2_tidy_samples, m2_filt_variant_paths, write_tsv)

# combine all mutect2 variants into a single list 
m2_somatic_vars <- dplyr::bind_rows(m2_tidy_samples) %>%
  dplyr::select(sample, SYMBOL, everything()) %>% 
  identity()

# save mutect2 variant df
m2_var_path <- "~/rb_pipeline/results/SNV/mutect2_variants.rds"

saveRDS(m2_somatic_vars, m2_var_path)


```

# 10. load custom filtered m2 vars

```{r m2_file_load10}

m2_var_path <- "~/rb_pipeline/results/SNV/mutect2_variants.rds"
m2_somatic_vars <- readRDS(m2_var_path)

# retrieve split list of m2 variants
m2_somatic_vars_list <- split_m2_somatic_vars(m2_somatic_vars)

purrr::map2(m2_somatic_vars_list, names(m2_somatic_vars_list), print_tl)

```


# 2. collate RB1 mutect2 variants for all samples

```{r m2_file_load13}

tidy_and_select_rb <- function(variants){
  browser()
  variants <- variants %>% 
    dplyr::filter(SYMBOL == "RB1" & !is.na(genename)) %>% 
    dplyr::select(one_of(c("sample", "seqnames", "start", "end", "AF.TUMOR")))
}

test0 <- 
  m2_somatic_vars_list[c("vc_T", "vc_CL")] %>% 
  purrr::map(tidy_and_select_rb) %>%
  dplyr::bind_rows() %>%
  identity()

# m2_somatic_vars_rb <- tidy_and_select_rb(m2_somatic_vars)
# 
# rb1_var_from_bam_igv <- read_csv("~/rb_pipeline/results/rb1_var_not_mutect2.csv")
# 
# 
# tidy_rb_variants <- rbindlist(list(m2_somatic_vars_rb, rb1_var_from_bam_igv), fill = T) %>% 
#   arrange(sample)
# 
# rb1_status_old <- read_csv("~/rb_pipeline/doc/rb1_status_old.csv") %>% 
#   dplyr::select(-c(somatic.snv.VAF, somatic.snv.AAchange, germline.snv.VAF, germline.snv.AAchange))
# 
# rb1_status <- full_join(tidy_rb_variants, rb1_status_old, by = c("sample" = "Sample.ID")) %>% 
#   dplyr::rename(chrom = seqnames)
# 
# rb1_status <- rb1_status[gtools::mixedorder(rb1_status$sample),]
# 
# write_csv(rb1_status, "~/rb_pipeline/doc/RB_exome_manuscript/rb1_status2.csv")


```

# 3. Reported Vision Center Variants

```{r m2_file_load14}

# reported m2 somatic vars
# filter by gatk filter and pathogenicity

m2_vc_somatic_vars <- bind_rows(m2_somatic_vars_list$vc_CL, m2_somatic_vars_list$vc_T)

reported_m2_vc_somatic_vars <- filter(m2_vc_somatic_vars, FILTER == "PASS") %>% 
  filter(path_meta_score >= 2) %>% 
  select(sample, SYMBOL, HGVSc, HGVSp, everything()) %>%
  dplyr::group_by(sample, seqnames, start, end) %>% # each variant has >1 protein product, need to filter each down to one
  dplyr::filter(row_number() == 1) %>%
  dplyr::filter(!is.na(SYMBOL)) %>% 
  dplyr::filter(sample != "31-T") %>% 
  identity() 

write_csv(reported_m2_vc_somatic_vars, "../doc/RB_exome_manuscript/SNV/reported_m2_vc_somatic_variants.csv")

```

# 4. Reported Reynolds Variants

```{r m2_file_load15}

# reported m2 somatic vars 
# filter by gatk filter and pathogenicity

m2_reynolds_somatic_vars <- m2_somatic_vars_list$reynolds_CL

reported_m2_reynolds_somatic_vars <- filter(m2_reynolds_somatic_vars, FILTER == "PASS") %>% 
    dplyr::mutate(path_meta_score = ifelse(SIFT_score >=0.5, 1, 0) +
                      ifelse(Polyphen2_score >=0.5, 1, 0) +
                      ifelse(MutationTaster_score >=0.5, 1, 0) +
                      ifelse(LRT_score >=0.5, 1, 0)) %>%
  filter(path_meta_score >= 2) %>% 
  dplyr::distinct(sample, seqnames, start, end, .keep_all = T) %>% 
  select(sample, SYMBOL, HGVSc, HGVSp, everything()) %>%
  dplyr::group_by(sample, seqnames, start, end, SYMBOL) %>% # each variant has >1 protein product, need to filter each down to one
  dplyr::filter(row_number() == 1) %>%
  dplyr::filter(!is.na(SYMBOL)) %>% 
  dplyr::filter(!is.na(HGVSc)) %>% 
  identity() 

write_csv(reported_m2_reynolds_somatic_vars, "../doc/RB_exome_manuscript/SNV/reported_m2_reynolds_somatic_variants.csv")

```


```{r }
test0 <- filter(m2_tidy_reynolds, FILTER == "PASS") %>% 
  dplyr::mutate(path_meta_score = ifelse(SIFT_score >=0.5, 1, 0) +
                      ifelse(Polyphen2_score >=0.5, 1, 0) +
                      ifelse(MutationTaster_score >=0.5, 1, 0) +
                      ifelse(LRT_score >=0.5, 1, 0)) %>%
  filter(path_meta_score >= 2) %>% 
  select(sample, SYMBOL, HGVSc, HGVSp, everything()) %>%
  dplyr::group_by(sample, seqnames, start, end, SYMBOL) %>% # each variant has >1 protein product, need to filter each down to one
  dplyr::filter(row_number() == 1) %>%
  dplyr::filter(!is.na(SYMBOL)) %>% 
  dplyr::filter(!is.na(HGVSc)) %>% 
  identity() 
```

# 5. load reported variants

```{r m2_file_load16}
reported_vars_path <- "../doc/RB_exome_manuscript/SNV/reported_m2_vc_somatic_variants.csv"
reported_m2_vc_somatic_vars <- read_csv(reported_vars_path)

reported_cl_vars <- dplyr::filter(reported_m2_vc_somatic_vars, grepl("CL", sample))
write_csv(reported_cl_vars, "SNV/clean_vc_cell_line_variants.csv")
print_tl(reported_cl_vars, "final m2 cl variants")

reported_t_vars <- dplyr::filter(reported_m2_vc_somatic_vars, grepl("T", sample))
write_csv(reported_t_vars, "SNV/clean_vc_tumor_variants.csv")
print_tl(reported_t_vars, "final m2 t variants")

#### load reported reynolds variants
reported_m2_reynolds_somatic_vars <- read_csv("../doc/RB_exome_manuscript/SNV/reported_m2_reynolds_somatic_variants.csv")

short_stat <- function(stat_function, vector){
  signif(stat_function(vector), digits = 2)
}

t_vars_by_sample <- as.numeric(table(reported_t_vars$sample))
glue::glue("{short_stat(mean, sample_df)} +/- {short_stat(sd, sample_df)} range {min(sample_df)}-{max(sample_df)}", sample_df = t_vars_by_sample)

cl_vars_by_sample <- as.numeric(table(reported_cl_vars$sample))
glue::glue("{short_stat(mean, sample_df)} +/- {short_stat(sd, sample_df)} range {min(sample_df)}-{max(sample_df)}", sample_df = cl_vars_by_sample)

```

# 11. plot average VAF per sample

```{r m2_file_load11}

ggplot(m2_somatic_vars, aes(x = AF.TUMOR, y = sample)) + geom_density_ridges()

```

# 12. box plot somatic variants per sample

```{r m2_file_load12}
vc_somatic_vars <- dplyr::filter(m2_somatic_vars, !grepl("[0-9]{3}-CL", sample)) %>% 
  ungroup() %>% 
  dplyr::count(sample) %>% 
  mutate(sample_type = ifelse(grepl("T", sample), "Tumor", "Cell Line"))

ggplot(vc_somatic_vars, aes(x = sample_type, y = n, label = sample)) + 
  scale_y_log10() + 
  geom_boxplot() + 
  geom_text_repel(color = "red", cex = 2.5)

```

